Gradual Soundness: Lessons from Static Python

### 2021-10-20 ecoop-2022/paper.tex
ecoop: more sec2 filling

There are a few restrictions on classes
and a few optional types that programmers may wish to use,
but otherwise Python programs can run via \SP{}.
%% TODO awk

@@

\Cref{fig:sp-example} (left) presents a first example program
that defines a function \code{f} and calls it from typed code.
The syntax is normal Python with type annotations.
Only the import statement makes special use of \SP{};
it imports a type \code{PyDict} that describes Python dictionaries.

Although this program is typed, a different untyped module is free
to import the function \code{f} and invoke it with any sort of
argument.
For this reason, the compiled version of \cref{fig:sp-example} must
include a few run-time casts:
\begin{itemize}
  \item
    The function \code{f} must check that its arguments match the \code{PyDict} domain annotation.
    Concretely, this check accepts any dictionary.
  \item
    Because the domain check does not validate the contents of an incoming dictionary,
    the dict access must check that its result matches the function codomain.
\end{itemize}
These checks ensure a kind of type soundness.

@@

@@

%% more to say about the style of migration in SP?

@@

\subsection{Restrictions}

No overloading.
At most one method per class and name.
Ensures that receiver type does not affect dispatch.

Overrides must be compatible, if done in static code.
What's the reason?

\subsection{Additional Options}

@@

@@

\section{Future Work}

>>>>

Almost all Python programs can run under \SP{} with no changes.
The few exceptions are due to restrictions that \SP{} puts
on classes to promote fast-running bytecode.

@@

\Cref{fig:sp-example} presents a first example program
that defines a function \code{f} and calls it from typed code.
The syntax is normal Python with type annotations.
Only the import statement makes special use of \SP{};
it imports a type \code{PyDict} that describes Python dictionaries.

Although this program is typed, a different untyped module is free
to import the function \code{f} and invoke it with any sort of
argument.
For this reason, the compiled version of \cref{fig:sp-example} must
include a few run-time casts:
\begin{itemize}
  \item
    The function \code{f} must check that its arguments match the \code{PyDict} domain annotation.
    Concretely, this check accepts any Python dictionary.
  \item
    Because the domain check does not validate the contents of an incoming dictionary,
    the dict access must check that its result matches the function codomain.
\end{itemize}
These checks ensure a kind of type soundness.

@@

[FILL]

@@

%% more to say about the style of migration in SP?
[FILL]

@@

\subsection{Virtual Method Tables}

\SP{} adds vtables to Python.
[FILL]

\subsection{Code Restrictions}

\SP{} imposes several restrictions on classes and objects
to support efficient compilation:

\begin{itemize}
  \item \textbf{First-order Classes}.
    In order to benefit from types, class definitions must be defined at the top level of a module.
    Nested classes are supported by the language, but have type \code{Dynamic} according
    to the type checker.

  \item \textbf{No overloading}.
    Methods cannot be overloaded; each combination of a class
    and a name can refer to at most one method.
    This ensures that method resolution depends only on the
    type of the receiver object.
    % resolution does NOT depend on result type (T doesn't matter in `x:T = o.m()`)
    % Q. does resolution depend on args?

  %\item \textbf{Single Inheritance}.
  %  Multiple inheritance is not supported.
  %  But, it might someday be supported.

  \item \textbf{Compatible method overrides}.
    Methods may be overridden, either by \SP{} classes or
    untyped Python classes.
    Overrides that appear in \emph{the same} \SP{} file as the class definition, however,
    must supply a method whose type is a static subtype of the previous method.
    In particular, a parent method that returns type \code{int} cannot be
    overridden by a child method that returns type \code{Dynamic}.
    %% TODO what does this enable ... optimizations in the same file?

  % NOTE __ (dunder) methods may be overridden e.g. __getattribute__,
  % but the normal field access syntax skips any overrides (o.f)

  \item \textbf{Compatible field overrides}.
    Similarly, object fields and class variables may be overridden
    only with subtype-compatible values.
    In fact, \SP{} compiles fields using Python slots
    declarations (\code{\_\_slots\_\_}).
    This declaration prevent new fields from arising at runtime,
    but saves space and speeds up field accesses.

\end{itemize}

\subsection{Type Restrictions}

Certain types are not supported:

\begin{itemize}
  \item \textbf{Parameterized Classes}.
    At present, \SP{} treats all class parameters as type \code{Dynamic}.
    The team plans to lift this restriction (\cref{s:future}).

  \item \textbf{Callable}.
    The type system cannot yet express first-class functions.
    The team plans to lift this restriction~(\cref{s:future}),
    but there is little internal pressure to do so.

  \item \textbf{Dynamic Checked Types}.
    A \code{CheckedDict} cannot be instantiated with \code{Dynamic}
    keys or values.
    [FILL] no desire, and makes runtime checks much simpler.

  \item \textbf{Recursive Types}.
    Sadly this is normal for Python type checkers.

\end{itemize}

\subsection{Additional Features}

@@

ctypes for primitives, unboxed representation

@@

\section{Future Work}
\label{s:future}
